---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  size = "footnotesize"
)
```

# harmBounds

The harmBounds package calculates stopping probabilities,
	defines stopping boundaries and generates plots 
	for safety monitoring using an event based approach.

## Installation

The package can be installed from [GitHub](https://github.com/):

```{r, eval=FALSE}
# install.packages("devtools")
remotes::install_github("dcr-unibe-ch/harmBounds")

# load package
library(harmBounds)
```

```{r, echo=FALSE, results="hide", warning=FALSE, message=FALSE}
devtools::load_all()
options(width = 200)
```

## Basic usage

### Stopping boundaries

```{r message=FALSE, warning=FALSE}
hb<-getHarmBound(nevents = seq(10, 100, by = 10), alpha_test = 0.025, pH0 = 0.5,
  maxevents = 150)
plot(hb)
```

### Operating characteristics

Stopping probabilities and expected number of events can be obtained for alternative scenarios.

```{r}
hb<-getHarmBound(nevents = seq(10, 100, by = 10), alpha_test = 0.025, 
  pH0 = 0.5, pH1 = c(0.6, 0.7, 0.8) , maxevents = 150)
plot(hb, which = "cum_stopping")
```

The above example only scratches the surface of what `harmBounds` can do. Check the 
vignette for further examples.
